---
title: "bgns: Biweight Graph and Network Statistics"
author: "Aditya Kshirsagar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bgns: Biweight Graph and Network Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

`bgns` provides biweight midcorrelation and exact bicor-based k-nearest-neighbor
KNN graph construction for dense and sparse numeric matrices. bgns builds graphs
from robust correlation structure, supporting workflows where similarity is
better represented by coordinated variation. The package supports dense matrices
and sparse `Matrix::dgCMatrix` inputs, and the tidy and KNN paths are panelized
so that a full dense similarity matrix does not need to be stored.

The package expects rows to be observations and columns to be the items being
compared, such as genes, cells, metacells, samples, or features.

## Installation

```{r installation, eval=FALSE}
install.packages("bgns")
```

## Dense bicor

```{r dense-bicor}
library(bgns)
set.seed(1)
x <- matrix(rnorm(240), nrow = 40, ncol = 6)
x[sample(length(x), 8)] <- NA_real_
colnames(x) <- paste0("item", seq_len(ncol(x)))

cmat <- bicor(x)
round(cmat[1:4, 1:4], 3)
```

For graph workflows, tidy output is often more useful than a dense matrix.

```{r tidy-bicor}
td <- bicor(x, tidy = TRUE, threshold = 0.1)
head(td)
```

## Exact bicor KNN graphs

`bicor_knn()` returns a directed edge table. `col1` is the target column, `col2`
is the selected neighbor, `val` is the bicor value, and `rank` is the within-target
rank.

```{r knn-bicor}
kn <- bicor_knn(x, knn = 2, threshold = -Inf)
kn
```

When `y` is supplied, neighbors are selected from columns of `x` for each target
column in `y`.

```{r bipartite-knn}
y <- matrix(rnorm(160), nrow = 40, ncol = 4)
colnames(y) <- paste0("target", seq_len(ncol(y)))
head(bicor_knn(x, y, knn = 2, bipartite_levels = "separate"))
```

Dense and sparse inputs can also be mixed in bipartite KNN calls.

```{r mixed-knn}
xs <- Matrix::rsparsematrix(40, 6, density = 0.3)
head(bicor_knn(xs, y, knn = 2, threshold = -Inf, bipartite_levels = "separate"))
```

## Missing values and thresholds

With `pairwise.complete.obs = TRUE`, each edge is computed on its finite overlap.
With `pairwise.complete.obs = FALSE`, KNN edges involving incomplete columns are
not eligible.

```{r pairwise-complete}
kn_pairwise <- bicor_knn(x, knn = 2, pairwise.complete.obs = TRUE)
kn_complete <- bicor_knn(x, knn = 2, pairwise.complete.obs = FALSE)

nrow(kn_pairwise)
nrow(kn_complete)
```

The `threshold` argument in `bicor_knn()` is a signed cutoff applied before
top-k selection; candidate bicor values must be strictly greater than the cutoff.
For example, `threshold = 0` keeps only positive bicor edges.

```{r threshold}
positive_edges <- bicor_knn(x, knn = 2, threshold = 0)
positive_edges
```

## Sparse matrices

Sparse inputs are accepted through `Matrix::dgCMatrix`; other sparse `Matrix` classes are coerced to `dgCMatrix` when possible.

```{r sparse}
xs <- Matrix::rsparsematrix(40, 6, density = 0.3)
colnames(xs) <- paste0("s", seq_len(ncol(xs)))

head(bicor(xs, tidy = TRUE, threshold = 0))
bicor_knn(xs, knn = 2, threshold = 0)
```

## Relationship to related packages

`WGCNA` provides a mature ecosystem for weighted correlation network analysis,
including module detection and network summaries. `bgns` is narrower: it focuses
on memory-efficient bicor kernels and exact bicor KNN graph construction.

`BiocNeighbors` provides nearest-neighbor search infrastructure for standard
geometric distances. `bgns` targets robust correlation-native KNN graphs where
coordinated variation is the similarity signal of interest.

## Runtime tuning

`BGNS_MEM_MB` controls the scratch-memory budget used for panelized computation.
When OpenMP is available, `BGNS_NUM_THREADS` controls the maximum number of
threads used by `bgns`; the default is at most two threads.

```{r runtime-tuning, eval=FALSE}
Sys.setenv(BGNS_MEM_MB = 512, BGNS_NUM_THREADS = 2)
```

## Internal execution engine

Version 0.4.2 keeps the public API focused on `bicor()` and `bicor_knn()` but uses a stronger internal execution layer. Long-running panel loops use conservative thread control through `BGNS_NUM_THREADS`, bounded panel sizes through `BGNS_MEM_MB`, deterministic top-k buffers, sparse `dgCMatrix` summaries, optional runtime-gated SIMD kernels where available, and interrupt checks at panel boundaries. These pieces are internal and do not change the mathematical definition of the bicor score.

## Session information

```{r session-info}
sessionInfo()
```
